---
title: "2016 IL Financial Contributions to Presidential Campaigns version 2"
author: "Sunny Jiang"
date: "November 10, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
```

***

## Data sources

This report explores a dataset of 2016 Illinois Financial Contributions to 
Presidential Campaigns, which including 250K records. 
For details please refer to this website: 
<http://classic.fec.gov/disclosurep/pnational.do>


```{r set working directory, bring in library. And clean up the data set}
setwd("C:/Users/jiangz4/Desktop/Udacity/Data Analyst/R Project")

library(ggplot2) #must load the ggplot package first
library(dplyr)
library(knitr)
library(treemap)
library(ggmap)
library(gridExtra)

data(zipcode)

# bring in the data set
il <- read.csv("IL_2016.csv", header = TRUE, fill = TRUE, sep = ",", 
               row.names = NULL)
colnames(il) <- c(colnames(il)[-1],"x")
il$x <- NULL

zipcode <- read.csv("zip_codes_states.csv", header = TRUE, fill = TRUE, sep = ",", 
               row.names = NULL)

# clean up the data to combine the multiple entries
il_clean <- il%>%
    group_by(cmte_id,cand_id,cand_nm,contbr_nm,contbr_city,contbr_st,
             contbr_zip,contbr_employer,contbr_occupation,contb_receipt_dt)%>%
    summarize(amount=sum(contb_receipt_amt))

# add clean columns (last name only, change date to right column type
# aggregate date to month-year, 5 digits zip code) and sort by amount
il_clean$cand_ln <- gsub(",.*$", "", il_clean$cand_nm) 
il_clean$contb_receipt_dt <- as.Date(il_clean$contb_receipt_dt, 
                                     format = "%d-%b-%y")
il_clean$contb_receipt_m_y <- as.Date(cut(il_clean$contb_receipt_dt,
                         breaks = "month"))
il_clean$zip <- substr(as.character(il_clean$contbr_zip),1,5)

il_clean <- il_clean[order(-il_clean$amount),]
```

```{r}
# group by candidate name, contirbution date and amount
il_groupby_cand_mth <- il_clean%>% 
    group_by(cand_nm,cand_ln,contb_receipt_m_y)%>%
    summarize(amount=sum(amount),count=n())
```

```{r}
# group by contributor, contirbution date and amount
il_groupby_contb <- il_clean%>%
    group_by(cmte_id,cand_id,cand_nm,contbr_nm,contbr_city,contbr_st,
             contb_receipt_m_y,zip,cand_ln)%>%
    summarize(amount=sum(amount))
```

```{r}
# summary table for contributor
il_groupby_contb_sum <- il_groupby_contb%>%
                        group_by(contbr_nm, zip)%>%
                        summarise(amount=sum(amount),times_of_contbr=n(),
                                  number_of_candi=length(unique(cand_id)))
il_groupby_contb_sum <- 
    il_groupby_contb_sum[order(-il_groupby_contb_sum$amount),]
```

```{r}
# summary table for contributor
il_groupby_contb_heat <- il_groupby_contb_sum%>%
    group_by(times_of_contbr, number_of_candi)%>%
    summarise(amount=sum(amount),count=n())
```

```{r}
# merge zipcode data to bring in lat and lon
il_groupby_contb_zip <- 
    merge(il_clean,zipcode,by,by.x='zip', by.y='zip_code')
```

```{r}
# summary table for treemap. group by county, city and zip
il_groupby_county_cand <- il_groupby_contb_zip%>%
    group_by(cand_ln, state, county, city, zip, latitude, longitude)%>%
    summarise(amount=sum(amount),count=n())
```


```{r}
# find the top contirbution counties
il_sum_county <- subset(il_groupby_county_cand, state == "IL") %>% 
    group_by(county) %>% summarise(amount = sum(amount), count = n())
il_sum_county <- il_sum_county[order(-il_sum_county$amount),]
```

***

## Data Explore and Analysis

### Univariate

Heatmap and box plot for contribution amount to decipt outliers
```{r}
# Heatmap and box plot for contribution amount to decipt outliers
grid.arrange( ggplot(aes(x=amount), 
                     data = il_clean) +
                  geom_histogram( bins = 30)+scale_x_log10() ,
              ggplot(aes(x=1, y=amount), 
                     data = il_clean) +
                  geom_boxplot( ) +scale_y_log10() , nrow =1)
```

To elimate the noise, we can plot only the ones within 1% to 99% percentile. 
We can see most contribution amounts are within $0 to $1000 range, 
after zoom in we can see most amount are actully less than $250.

```{r}
# contirbution amount by contributor histogram with quantile (0.01 to 0.99)
ggplot(data=il_groupby_contb, aes(amount)) + 
    geom_histogram(binwidth = 20) +
    xlim(quantile(il_groupby_contb$amount,0.01),
         quantile(il_groupby_contb$amount,0.99))
    
# contirbution amount by contributor histogram with 0 to 1000
ggplot(data=il_groupby_contb, aes(amount)) + 
    geom_histogram(binwidth = 20) +    
    xlim(0,1000)
```


From the the first boxplot we can see, this dataset has a few data points that
incorrectly mapped to Illinois. And when plot the count by county for IL. Cook
county contirbutes most of the contributions.

```{r}
# plot contribution amount by state
ggplot(aes(x=state,y=amount), data = il_groupby_contb_zip) + 
    geom_boxplot()

# plot contribution amount by county for IL
ggplot(aes(x=county), data = subset(il_groupby_contb_zip,state=="IL")) +
    geom_histogram(stat="count") +
    coord_flip()
```


From two bar charts we can see most people only contributed less than 3 times.
And only supported 1 candidate.

```{r}

# number of contirbutions made by each contributor histogram 
ggplot(data = il_groupby_contb_sum, aes(times_of_contbr)) + 
    geom_histogram(binwidth = 1)

# number of candidate supported by each contributor histogram 
ggplot(data = il_groupby_contb_sum, aes(number_of_candi)) + 
    geom_histogram(binwidth = 1)

```

***

### Bivariate

Take a closer look through the heatmap: 55% of people only 
contributed once to one candidate and covers 41% of total contribution 
amount in IL.


```{r}

# number of contirbutions and candidate supported heatmap
ggplot(data = il_groupby_contb_heat, 
       aes(x = number_of_candi, y = times_of_contbr)) +
    geom_tile(aes(fill = log(count)),colour = "white") +
    geom_text(aes(label = count)) +
    scale_fill_gradient(low = "white", high = "steelblue") +
    scale_y_continuous(breaks = c(1:20,1))
```

Contribution amount distribution for each candidate. Cruz has the highest
contribution amount.

```{r}
# contribution amount distribution for each candidate
ggplot(data=il_clean,
        aes(y=amount, x=cand_ln )) +
   geom_boxplot(  ) +
      theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

From these two bar charts, we can see **Clinton**, **Sanders**, **Cruz** and 
**Trump** are the top 4 condidates by both contribution count and amount

```{r pressure, echo=FALSE}
# plot by candidate name and contribution count
il_groupby_cand_plot1 <- ggplot(data=il_groupby_cand_mth, 
                                aes(x=reorder(cand_nm, count), y=count)) + 
    geom_bar(stat = "identity") +
    coord_flip() +
    xlab("Candidate Name") + ylab("Contribution Count")

# plot by candidate name and contribution amount
il_groupby_cand_plot2 <- ggplot(data=il_groupby_cand_mth, 
                                aes(x=reorder(cand_nm, amount), y=amount)) + 
    geom_bar(stat = "identity") +
    coord_flip() +
    xlab("Candidate Name") + ylab("Contribution Amount")

grid.arrange(il_groupby_cand_plot1, il_groupby_cand_plot2, ncol=2)
```

Contribution amount distribution for each county, this chart is too busy to see
anything. So I tried treemap.

```{r}
# contribution amount distribution for each county
ggplot(data=il_groupby_contb_zip,
        aes(y=amount, x=county )) +
   geom_boxplot(  ) +
      theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

From these two treemap charts, we can see Cook County has the highest
contribution amount. When I broke down to City and zipcode level for Cook 
County, Chicago is the top 1. Size corresponds to Amount and Color corresponds
to Count.

```{r , echo=FALSE}
# plot by county and contribution amount
treemap(subset(il_groupby_county_cand,amount > 0),
        index = c("county"),
        vSize = "amount",
        vColor = "count",
        type = "value",
        format.legend = list(scientific = TRUE, big.mark = " "))


treemap(subset(il_groupby_county_cand,county == "Cook"),
        index = c("city","zip"),
        vSize = "amount",
        vColor = "count",
        type = "value",
        format.legend = list(scientific = TRUE, big.mark = " "))
```


***

### Multivariate

For our top 4 candidates:

**Bernie Sanders** announced that he plan to seek the Democratic Party's 
nomination for President of the United States on May 26, 2015. On June 14, 2016, 
following the final primary election, Clinton became the presumptive Democratic 
nominee. As we can see from the time series chart, contribution made for Sanders
were between May 2015 to July 2016, peaked on March 2016. Matches his campaign 
timeline.

**Hillary Clinton** announced her presidential campaign on April 12, 2015. 
Nominated as the presumptive nominee of the Demoratic Party on July 26, 2016. 
Conceded the election to Republican Donald Trump on November 9. 
The contribution amount to her was the highest all the time, expect for 
exceeded by Sanders on March and April 2016.

**Ted Cruz** announced his presidential campaign on March 23, 2015. 
And suspended his campaign late April 2016. The contribution amount to him were 
higher than to Trump until he suspended his campaign.

**Donald Trump** formally launched his campaign on June 16, 2015. 
On May 3, 2016 Trump became the presumptive nominee of the Republican Party, 
and elected as president on November 8, 2016. From the chart we can see, 
contribution to him was really low at the beginning and started to take off 
on May 2016 after he became the presumptive nominee of Republican Party.


```{r}
# plot by candidate name and contribution amount over time
ggplot(data=subset(il_groupby_cand_mth, cand_ln %in% 
                       c("Trump", "Sanders", "Clinton", "Cruz")), 
       aes(x=contb_receipt_m_y, y=amount, color=cand_ln)) + 
    geom_line() +
    scale_x_date() + xlab("") + ylab("Total Contribution Amount")
```


Then study the distirbution for top 4 candidates with boxplot. 
Sanders has the lowest median contribution amount around $60. We can see some 
darker lines at $10, $30 and $100 which means a lot of contribution at 
those amounts. While Clinton has the hightest median contribution amount around 
$100, and have a lot of contributed at $50, $100, $500 and $2500. 
The highest contribution amount is $27000, and contributed to Cruz.

```{r}
# contirbution amount by contributor boxplot with log
ggplot(data=subset(il_groupby_contb, cand_ln %in% 
                       c("Trump", "Sanders", "Clinton","Cruz")), 
       aes(x=cand_ln, y=amount)) + 
    geom_jitter(alpha=1/50) +
    geom_boxplot() +
    scale_y_log10(breaks = c(1,5,10,20,50,70,100,200,300,
                             500,1000,2000,3000,5000,10000,20000))
```

The first treemap shows that for Cook County, most of the contribtuion went to
Clinton, then Sanders. From the second one, we can see Clinton had most of all 
the contribtutions. Trump and Sanders had similar amount.

```{r}
treemap(subset(il_groupby_county_cand,amount>0 & cand_ln %in% 
                       c("Trump", "Sanders", "Clinton", "Cruz")),
        index = c("county","cand_ln"),
        vSize = "amount",
        vColor = "amount",
        type = "value",
        format.legend = list(scientific = TRUE, big.mark = " "))
        
treemap(subset(il_groupby_county_cand,amount>0 & cand_ln %in% 
                       c("Trump", "Sanders", "Clinton", "Cruz")),
        index = c("cand_ln","county"),
        vSize = "amount",
        vColor = "amount",
        type = "value",
        format.legend = list(scientific = TRUE, big.mark = " "))

```

I picked the top 10 counties with the highest contribution amount. And used 
boxplot and barchart to see the distiribtution for top 4 candidates.
Contritubiton to Clinton from Cook county really dominated it. But in some
counties like Peoria, the contirbution to Trump was higher.

```{r}
head(il_sum_county,10)
ggplot(data=subset(il_groupby_county_cand,amount>0 & county %in% 
                       c("Cook", "Lake", "Du Page", "Kane", "Will", "Champaign",
                       "Mchenry", "Madison", "Peoria", "Winnebago") & 
                       cand_ln %in% c("Trump", "Sanders", "Clinton", "Cruz")),
        aes(y=amount, x=county  , fill = cand_ln)) +
   geom_boxplot(  ) +
   scale_y_log10() +
   coord_flip()
   
ggplot(data=subset(il_groupby_county_cand,amount>0 & county %in% 
                       c("Cook", "Lake", "Du Page", "Kane", "Will", "Champaign",
                       "Mchenry", "Madison", "Peoria", "Winnebago") & 
                       cand_ln %in% c("Trump", "Sanders", "Clinton", "Cruz")),
        aes(y=amount, x=county  , fill = cand_ln)) +
   geom_bar(stat="identity",position="dodge") +
   scale_fill_discrete(name="cand_ln") +
   coord_flip()
```



Plot the contribution amount by zipcode. Can't really find a lot information 
except we can identify the big cities like Chicago, Peoria, Springfield...

```{r}
il_groupby_contb_zip %>% group_by(state) %>% summarize(n())
# plot by zipcode
ggplot(data = subset(il_groupby_contb_zip, state == "IL"), 
       aes(x = longitude, y = latitude, color = log(amount))) +
    geom_point() 
```

***

## Final Plots and Summary

### Plot One

For 2016 Presidency Campaign, Sanders, Clinton, Cruz and Trump were the top 4 
candidates by contribution amount and count. This shows the contirubtion amount 
for these 4 leading candidates overtime. Clint has the highest amount all the 
time except for exceeded by Sanders on March and April 2016. This might be able
to indicate her winning of Illinois electoral votes.

**Bernie Sanders** announced that he plan to seek the Democratic Party's 
nomination for President of the United States on May 26, 2015. On June 14, 2016, 
following the final primary election, Clinton became the presumptive Democratic 
nominee. As we can see from the time series chart, contribution made for Sanders
were between May 2015 to July 2016, peaked on March 2016. Matches his campaign 
timeline.

**Hillary Clinton** announced her presidential campaign on April 12, 2015. 
Nominated as the presumptive nominee of the Demoratic Party on July 26, 2016. 
Conceded the election to Republican Donald Trump on November 9. 
The contribution amount to her was the highest all the time, expect for 
exceeded by Sanders on March and April 2016.

**Ted Cruz** announced his presidential campaign on March 23, 2015. 
And suspended his campaign late April 2016. The contribution amount to him were 
higher than to Trump until he suspended his campaign.

**Donald Trump** formally launched his campaign on June 16, 2015. 
On May 3, 2016 Trump became the presumptive nominee of the Republican Party, 
and elected as president on November 8, 2016. From the chart we can see, 
contribution to him was really low at the beginning and started to take off 
on May 2016 after he became the presumptive nominee of Republican Party.


```{r}
# plot by candidate name and contribution amount over time
ggplot(data=subset(il_groupby_cand_mth, cand_ln %in% 
                       c("Trump", "Sanders", "Clinton", "Cruz")), 
       aes(x=contb_receipt_m_y, y=amount, color=cand_ln)) + 
    geom_line() +
    scale_x_date() + xlab("") + ylab("Total Contribution Amount [USD]") +
    labs(title = "Total Contribution Amount for Leading Candidates Overtime") +
    scale_color_discrete(name="Candidate Last Name")
```

### Plot Two

Then study the distirbution for top 4 candidates with boxplot. 
Sanders has the lowest median contribution amount around $60. We can see some 
darker lines at $10, $30 and $100 which means a lot of contribution at 
those amounts. While Clinton has the hightest median contribution amount around 
$100, and have a lot of contributed at $50, $100, $500 and $2500. 
The highest contribution amount is $27000, and contributed to Cruz.

```{r}
# contirbution amount by contributor boxplot with log
ggplot(data=subset(il_groupby_contb, cand_ln %in% 
                       c("Trump", "Sanders", "Clinton","Cruz")), 
       aes(x=cand_ln, y=amount)) + 
    geom_jitter(alpha=1/50) +
    geom_boxplot() +
    scale_y_log10(breaks = c(1,5,10,20,50,70,100,200,300,
                             500,1000,2000,3000,5000,10000,20000)) +
    xlab("Candidate") + ylab("Contribution Amount [USD]") +
    labs(title = "Contribution Amount distribution by Leading Candidates")
```

### Plot Three

55% of people only contributed once to one candidate and covers 41% of total 
contribution amount in IL. This tells us for most people making contirbution was
a one time only thing through a campaign event. Less amount of people 
made contirbution repetitively or supported multiple candidates over time

```{r}
# number of contirbutions and candidate supported heatmap
ggplot(data = il_groupby_contb_heat, 
       aes(x = number_of_candi, y = times_of_contbr)) +
    geom_tile(aes(fill = log(count)),colour = "white", show.legend = TRUE) +
    geom_text(aes(label = count), show.legend = TRUE) +
    scale_fill_gradient(name = "Contirbution Counts",low = "lightblue", 
                        high = "steelblue") +
    scale_y_continuous(breaks = c(1:20,1)) +
    xlab("Number of Candidate Supported") + 
    ylab("Number of Times that made Contribution") +
    labs(title = "Contribution Frequency and Candidates Supported") +
    theme_classic()
```

***

## Reflection
* Since I'm living in Peoira Illinois, I choosed Illinois data for this study.
* I know Clint won the Illinois electoral votes, so I assumed she got most of 
the contribution. Which turns out true.
* When I tried to map the location by zip code, I noticed that there are a few 
records that actually outside Illinois. For future exploration I should 
eliminate those records up front.
* Another idea is group the zip code by county and compare with the final 
results. So we can see in detail whether the contribution amount affects the 
final results.


